---
title: "Quiz 3"
author: "KunZhu"
date: "December 27, 2015"
output: html_document
---

### Question 1
Consider the `mtcars` data set. Fit a model with mpg as the outcome that includes number of cylinders as a factor variable and weight as confounder. Give the adjusted estimate for the expected change in mpg comparing 8 cylinders to 4.

```{r}
fit <- lm(mpg ~ factor(cyl) + wt, data = mtcars)
summary(fit)$coefficient
```

### Question 2
Consider the `mtcars` data set. Fit a model with mpg as the outcome that includes number of cylinders as a factor variable and weight as a possible confounding variable. Compare the effect of 8 versus 4 cylinders on mpg for the adjusted and unadjusted by weight models. Here, adjusted means including the weight variable as a term in the regression model and unadjusted means the model without weight included. What can be said about the effect comparing 8 and 4 cylinders after looking at models with and without weight included?.
```{r}
fitua <- lm(mpg ~ factor(cyl), data = mtcars)
summary(fit)$coefficient
summary(fitua)$coefficient
```

**Solution:**Holding weight constant, cylinder appears to have less of an impact on mpg than if weight is disregarded.

### Question 3
Consider the `mtcars` data set. Fit a model with mpg as the outcome that considers number of cylinders as a factor variable and weight as confounder. Now fit a second model with mpg as the outcome model that considers the interaction between number of cylinders (as a factor variable) and weight. Give the P-value for the likelihood ratio test comparing the two models and suggest a model using 0.05 as a type I error rate significance benchmark.
```{r}
fitua <- lm(mpg ~ factor(cyl), data = mtcars)
summary(fit)$coefficient
summary(fitua)$coefficient
```
**Solution:** The P-value is larger than 0.05. So, according to our criterion, we would fail to reject, which suggests that the interaction terms may not be necessary.

### Question 4 
Consider the `mtcars` data set. Fit a model with mpg as the outcome that includes number of cylinders as a factor variable and weight inlcuded in the model as
```{r}
lm(mpg ~ I(wt * 0.5) + factor(cyl), data = mtcars)
```
**Solution:** Since the unit of ``(wt * 0.5)`` is ```(lb/2000)```, and one (short) ton is 2000 lbs, the wt coefficient is interpretted as the estimated expected change in MPG per one ton increase in weight for a specific number of cylinders (4, 6, 8).

### Question 5 
Consider the following data set
```{r}
x <- c(0.586, 0.166, -0.042, -0.614, 11.72)
y <- c(0.549, -0.026, -0.127, -0.751, 1.344)
```
Give the hat diagonal for the most influential point

```{r}
x <- c(0.586, 0.166, -0.042, -0.614, 11.72)
y <- c(0.549, -0.026, -0.127, -0.751, 1.344)
fitxy <- lm(y ~ x)
hatvalues(fitxy)
```

### Question 6 
Consider the following data set
```{r}
x <- c(0.586, 0.166, -0.042, -0.614, 11.72)
y <- c(0.549, -0.026, -0.127, -0.751, 1.344)
fit6 <- lm(y ~ x)
dfbetas(fit6)[, 2]
```

### Question 7 
Consider a regression relationship between Y and X with and without adjustment for a third variable Z. Which of the following is true about comparing the regression coefficient between Y and X with and without adjustment for Z.
Solutions

**Solution:** It is possible for the coefficient to reverse sign after adjustment. For example, it can be strongly significant and positive before adjustment and strongly significant and negative after adjustment.
