---
title: "Reproducible Research: Peer Assessment 2"
output: 
  html_document:
    keep_md: true
---
Created by Kun Zhu on Aug 23rd, 2015

## Severe Weather Event Impact on Society 
### - Population Health and Economic Consequences, the U.S., 1996 - 2011

### Synopsis
This report provides a high level summary of the damage caused by severe weather events in the U.S. 
Specially, the number of injuries and fatalities and damage (in $) on properties and crops are examined. The original data is from the NAtional Oceanic and Atmospheric Administration and includes records from 1950 to 2011. Records from the early years, i.e., before 1996, are incomplete, therefore, this report focus on the data between 1996 and 2011. This study suggests excessive heat,  tornado, and flood are the major threats to population health, while flood, drought, and hurricane/typhoon have the greatest economic consequences.

### Data processing
* Load and overview the orignal dataset
```{r}
library(ggplot2)
library(gridExtra)
cache = TRUE
stormData <- read.csv("repdata_data_StormData.csv")
stormData$year <- as.numeric(format(as.Date(stormData$BGN_DATE, format = "%m/%d/%Y %H:%M:%S"), "%Y"))
hist(stormData$year, breaks = 60, col= "steelblue", xlab="Year", ylab="Num of Records", main = "Number of Storm Event Record, 1950-2011")
```

As the dataset from the recent years is more complete, this report focuses on statistics from the recent 15 years, i.e., 1996-2011.

* Subset data from 1996
```{r}
cache = TRUE
subsetData <- subset(stormData, year > 1995)
```
The property damage and crop damage data are updated with their corresponding unit multipliers from PROPDMGEXP and CROPDMGEXP columns.

* List the units from PROPDMGEXP and convert them to numerics
```{r}
cache = TRUE
unique(subsetData$PROPDMGEXP)
subsetData[subsetData$PROPDMGEXP == "H", ]$PROPDMG <- subsetData[subsetData$PROPDMGEXP == "H", ]$PROPDMG * 10^2
subsetData[subsetData$PROPDMGEXP == "h", ]$PROPDMG <- subsetData[subsetData$PROPDMGEXP == "h", ]$PROPDMG * 10^2
subsetData[subsetData$PROPDMGEXP == "K", ]$PROPDMG <- subsetData[subsetData$PROPDMGEXP == "K", ]$PROPDMG * 10^3
subsetData[subsetData$PROPDMGEXP == "M", ]$PROPDMG <- subsetData[subsetData$PROPDMGEXP == "M", ]$PROPDMG * 10^6
subsetData[subsetData$PROPDMGEXP == "m", ]$PROPDMG <- subsetData[subsetData$PROPDMGEXP == "m", ]$PROPDMG * 10^6
subsetData[subsetData$PROPDMGEXP == "B", ]$PROPDMG <- subsetData[subsetData$PROPDMGEXP == "B", ]$PROPDMG * 10^9
```
* List the units from CROPDMGEXP and convert them to numerics
```{r}
cache = TRUE
unique(subsetData$CROPDMGEXP)
subsetData[subsetData$CROPDMGEXP == "K", ]$CROPDMG <- subsetData[subsetData$CROPDMGEXP == "K", ]$CROPDMG * 10^3
subsetData[subsetData$CROPDMGEXP == "k", ]$CROPDMG <- subsetData[subsetData$CROPDMGEXP == "k", ]$CROPDMG * 10^3
subsetData[subsetData$CROPDMGEXP == "M", ]$CROPDMG <- subsetData[subsetData$CROPDMGEXP == "M", ]$CROPDMG * 10^6
subsetData[subsetData$CROPDMGEXP == "m", ]$CROPDMG <- subsetData[subsetData$CROPDMGEXP == "m", ]$CROPDMG * 10^6
subsetData[subsetData$CROPDMGEXP == "B", ]$CROPDMG <- subsetData[subsetData$CROPDMGEXP == "B", ]$CROPDMG * 10^9
```

### Severe weather event impact on population health
* Top 10 severe weather events for fatalities
```{r}
# Aggregate, subset and sort the dataset 
fatalitiesData <- aggregate(subsetData$FATALITIES, by=list(subsetData$EVTYPE), FUN=sum, na.rm=T)
names(fatalitiesData)[1] <- "Event"
names(fatalitiesData)[2] <- "Fatality"
sortedFatalities <-head(fatalitiesData[order(-fatalitiesData$Fatality),], 10)
sortedFatalities$Rank <-1:10
sortedFatalities <- sortedFatalities[c(3,1,2)]
sortedFatalities
```
* Top 10 severe weather events for injuries
```{r}
# Aggregate, subset and sort the dataset 
injuriesData <- aggregate(subsetData$INJURIES, by=list(subsetData$EVTYPE), FUN=sum, na.rm=T)
names(injuriesData)[1] <- "Event"
names(injuriesData)[2] <- "Injury"
sortedInjuries <- head(injuriesData[order(-injuriesData$Injury),], 10)
sortedInjuries$Rank <-1:10
sortedInjuries <- sortedInjuries[c(3,1,2)]
sortedInjuries
```
```{r fig.width=11, fig.height=6}
cache = TRUE
sortedInjuries <- transform(sortedInjuries, Event = reorder(Event, -Injury))
injuryPlot <- ggplot(sortedInjuries, aes(x=Event, y=Injury, fill=Event)) + 
  geom_bar(stat = "identity", width=.8) + 
  theme(axis.text.x=element_blank()) + 
  xlab("Severe Weather Type") + 
  ylab("Num of Injuries") +  
  ggtitle("Total Injuries Caused by Top 10 Severe Weather Events \n in the U.S., 1996 - 2011")

sortedFatalities <- transform(sortedFatalities, Event = reorder(Event, -Fatality))
fatalityPlot <- ggplot(sortedFatalities, aes(x=Event, y=Fatality, fill=Event)) + 
  geom_bar(stat = "identity", width=.8) + 
  theme(axis.text.x=element_blank()) + 
  xlab("Severe Weather Type") + 
  ylab("Num of Fatalities") +  
  ggtitle("Total Fatalities Caused by Top 10 Severe Weather Events \n in the U.S., 1996 - 2011")

grid.arrange(fatalityPlot, injuryPlot, nrow = 2)

```

### Damage on property and crop due to sever weather events
* Top 10 severe weather events for property damage
```{r}
# Aggregate, subset and sort the dataset 
propDmg <- aggregate(subsetData$PROPDMG, by=list(subsetData$EVTYPE), FUN=sum, na.rm=T)
names(propDmg)[1] <- "Event"
names(propDmg)[2] <- "PropDMG"
sortedpropDmg <-head(propDmg[order(-propDmg$PropDMG),], 10)
sortedpropDmg$Rank <-1:10
sortedpropDmg <- sortedpropDmg[c(3,1,2)]
sortedpropDmg
```
* Top 10 severe weather events for crop damage
```{r}
# Aggregate, subset and sort the dataset 
cropDmg <- aggregate(subsetData$CROPDMG, by=list(subsetData$EVTYPE), FUN=sum, na.rm=T)
names(cropDmg)[1] <- "Event"
names(cropDmg)[2] <- "CropDMG"
sortedcropDmg <-head(cropDmg[order(-cropDmg$CropDMG),], 10)
sortedcropDmg$Rank <-1:10
sortedcropDmg <- sortedcropDmg[c(3,1,2)]
sortedcropDmg 
```
```{r fig.width=11, fig.height=6}
sortedpropDmg <- transform(sortedpropDmg, Event = reorder(Event, -PropDMG))
propPlot <- ggplot(sortedpropDmg, aes(x=Event, y=PropDMG, fill=Event)) + 
  geom_bar(stat = "identity", width=.8) + 
  theme(axis.text.x=element_blank()) + 
  xlab("Severe Weather Type") + 
  ylab("Property Damage") +  
  ggtitle("Property Damage Caused by Top 10 Severe Weather Events \n in the U.S. from 1996 - 2011")

sortedcropDmg <- transform(sortedcropDmg, Event = reorder(Event, -CropDMG))
cropPlot <- ggplot(sortedcropDmg, aes(x=Event, y=CropDMG, fill=Event)) + 
  geom_bar(stat = "identity", width=.8) + 
  theme(axis.text.x=element_blank()) + 
  xlab("Severe Weather Type") + 
  ylab("Crop Damage") +  
  ggtitle("Crop Damage Caused by Top 10 Severe Weather Events \n in the U.S. from 1996 - 2011")

grid.arrange(propPlot, cropPlot, nrow = 2)
```

### Conclusion
This study suggests that excessive heat, tornato, and flood place the greatest threats to population healty in the U.S. From a economic point of view, flood, hurrican/typhoon, and droughts are the major causes of property and crop damage. 

